\documentclass[../ops.tex]{subfiles}

\begin{document}
        \subsection{Genesis}
        In the beginning MIT, AT\&T Bell Labs and General Electric worked on
        Multics. Now the Multics was formless and unfinished, darkness was over
        the surface of its shell and the spirit of Unix was hovering over the
        codebase.

        And Bell Labs said, ``Let us withdraw'' as they were not satisfied with
        the progress. And on the third day an employee of Bell Labs, Ken
        Thompson and co-worker of his, Dennis Ritchie discovered a PDP-7
        minicomputer. And they saw that the PDP was good and they decided to
        try again.
        
        \subsection{To the point}
        You ever seen \emph{The C Programming Language}? The book I mean.
        Remember who wrote it? Now take a peek at second paragraph of previous
        subsection. Rings a bell?

        First Unix was created at AT\&T Bell Labs in 1969. It was written in
        assembly language for a PDP-7 machine and soon was ported to PDP-11.
        It was a simplification of a project that K\&R worked on before
        -~Multics.

        Unix had a hierarchical file system \emph{(take that 1983 MS-DOS 2.0)},
        an assembler, a text editor and a shell. In 1973 Version 4 Unix was
        rewritten in C, an imperative high-level programming language created
        one year earlier by Dennis Ritchie and Ken Thompson for the purpose of
        making working with Unix easier, and the system itself more portable.

        Since Bell Labs was forbidden from entering any other business than
        communications services due to a 1956 decree in an antitrust case of its
        parent organization, Unix code was distributed to universities and other
        research entities\footnote{so called \emph{Research Unix}}.

        One of the places where this code landed was the Computer Systems
        Research Group at the university of California, Berkeley. Folk there
        worked hard on Unix, adding a lot of new stuff such as the {\tt vi} text
        editor, {\tt curses} library or integrating TCP/IP stack into it.

        This resulted in divergence between the AT\&T and Berkeley unices, the
        latter known as the Berkeley Software Distribution, or BSD for short.
        This funny little series of events led to a silly little thingy called
        {\bf Unix wars}, where AT\&T, Microsoft, IBM and BSD all fought over who
        has the best Unix and whose Unix should become the standard. The
        contenders were:
        \begin{itemize}
                \item UNIX System V by AT\&T
                \item Xenix by Microsoft
                \item PC/IX by IBM
                \item 4.2BSD by Berkeley
        \end{itemize}

        This is actually really fun because there will be one more unix-related
        war we will describe soon.
        Meanwhile: the result of the Unix wars was that AT\&T sued Berkeley,
        they won and BSD had to be cleansed of proprietary AT\&T code. And thus
        4.4BSD-Lite was created. Remember this one, we will get back to it soon.

        \subsection{Rise of the GNU}
        In 1983, out of nowhere a new actor comes onto the scene -~Richard
        Stallman of MIT. He starts what will be known as the GNU project. His
        aim? \emph{A complete Unix-compatible software system} composed entirely
        of free software. Software that everyone can use, share and modify, with
        open access to the source code.

        To make his dream true he retires from MIT for full independence, starts
        the Free Software Foundation (FSF) and creates GNU GPL copyleft license
        in order to protect aforementioned rights. By early 90's most of GNU
        software is ready, however the kernel is still incomplete.

        In the meantime American-Dutch computer scientist Andrew S. Tanenbaum
        writes a textbook on operating system while working at Vrije
        Universeiteit in Amsterdam. In order to exemplify things he writes about
        he creates MINIX operating system and makes its source code available
        for study and research to universities.

        This is important for the GNU timeline/multiverse because at this time,
        somewhere in Finland a student of University of Helsinki, Linus Torvalds
        reads Tanenbaum's book and works on MINIX. He starts a hobbyistic
        project and tries to write an operating system kernel. \emph{``just a
        hobby, won't be big and professional like gnu''}.

        And now the plotlines join, because in 1991 he releases first version of
        the Linux kernel. Inspired by a speech of Stallman that his friend took
        him to, he licenses his work under GNU GPLv2.

        Now check it out: what did the GNU project lack? {\bf A kernel!} Guess
        what, people start joining GNU with Linux, creating what is known as
        GNU/Linux distributions.

        \subsection{Summary}
        The story I just abstracted, real-life consequences follow that touch us
        in present day, present time:
        \begin{itemize}
                \item *nix systems are divided into three branches, based on
                        three of the most influential projects: the UNIX System
                        V (SysV) branch, focused on stability and
                        commerciability, BSD-licensed systems based on
                        4.4BSD-Lite, with a long tradition of hackers flexing
                        with their skills and showing innovation, and GNU/Linux
                        distros which are free, open source and written from
                        scratch with no direct inheritance of any OG Unix code.
                \item There also exists MINIX, which nowadays is used mostly
                        for educational purposes, though it has its industrial
                        uses as well, for example its modified version runs on
                        the Intel Management Engine
                \item This entire mess had to be standardized somehow. It was,
                        once by IEEE (POSIX standard) and once by The Open Group
                        (SUS\footnote{im going insane}). Nowadays those two
                        standards are more or less equivalent, but we usually
                        refer to POSIX if we want to make sure something is
                        portable across all unices.
        \end{itemize}
\end{document}
